---
title: "SAINTS - Sample size / power calculation - seroprevalance and seroincidence"
author: "Helen Dale"
date: "2022-09-05"
output:
  pdf_document: default
    code_folding: hide
classoption: landscape

---

```{r setup, results='hide', message=F, warning=F}
# options(width = 60)
knitr::opts_chunk$set(echo = TRUE, warning=F, message=F, results=F, fig.width=16, fig.height=9)

library(png)
library(tidyverse)
library(ggplot2)
library(pwr)
library(RColorBrewer)
library(longpower)

## Simulate dataset for serosurveys

x<-seq(0,1,by=0.01)
y<-rep(NA,length(x))
yl<-rep(NA,length(x))
yll<-rep(NA,length(x))
ylu<-rep(NA,length(x))
yu<-rep(NA,length(x))
yul<-rep(NA,length(x))
yuu<-rep(NA,length(x))
moe<-rep(NA,length(x))
moel<-rep(NA,length(x))
moeu<-rep(NA,length(x))
moeAlt1<-rep(NA,length(x))
moeAlt2<-rep(NA,length(x))

n<-200
nAlt1<-100
nAlt2<-400
nSim<-1e3

for(i in 1:length(x)){
    print(paste(sep="/",i,length(x)))
    tmp_y<-rep(NA,nSim)
    tmp_yu<-rep(NA,nSim)
    tmp_yl<-rep(NA,nSim)
    tmp_moeAlt1<-rep(NA,nSim)
    tmp_moeAlt2<-rep(NA,nSim)
    
    for(j in 1:nSim){
        samp<-rbinom(n=1,size=n,p=x[i])
        binEst<-binom.test(x=samp,n=n)
        tmp_y[j]<-binEst$estimate
        tmp_yl[j]<-binEst$conf.int[1]
        tmp_yu[j]<-binEst$conf.int[2]

        samp<-rbinom(n=1,size=nAlt1,p=x[i])
        binEst<-binom.test(x=samp,n=nAlt1)
        tmp_moeAlt1[j]<-(binEst$conf.int[2]-binEst$conf.int[1])/2
        samp<-rbinom(n=1,size=n,p=x[i])
        binEst<-binom.test(x=samp,n=nAlt2)
        tmp_moeAlt2[j]<-(binEst$conf.int[2]-binEst$conf.int[1])/2
    }
    
    tmp_moe<-(tmp_yu-tmp_yl)/2

    y[i]<-mean(tmp_y)
    yl[i]<-quantile(tmp_y,probs=0.025); yll[i]<-quantile(tmp_yl,probs=0.025); ylu[i]<-quantile(tmp_yl,probs=0.975)
    yu[i]<-quantile(tmp_y,probs=0.975); yul[i]<-quantile(tmp_yu,probs=0.025); yuu[i]<-quantile(tmp_yu,probs=0.975)
    moe[i]<-mean(tmp_moe); moel[i]<-quantile(tmp_moe,probs=0.025); moeu[i]<-quantile(tmp_moe,probs=0.975)
    moeAlt1[i]<-mean(tmp_moeAlt1)
    moeAlt2[i]<-mean(tmp_moeAlt2)
}

```

## 1. Seroprevalance for serosurveys

### Method
Data simulated (100,000 simulations) based on a binomial distribution
200 children per 1 year age-band

Acceptable margin of error and confidence interval for estimation across all seroprevalences for 1000 single samples (200 in each of 5 age bands), as a minimum sample size for each site in Malawi, Kenya, Burkina Faso and Ghana.

```{r}
par(mfrow=c(1,2))

plot(x,y,xlab="prevalence",ylab="estimated prevalence",col="black",type="l",lwd=2,asp=1,main="n=200")

tmpCol<-col2rgb("steelblue")
polygon(x=c(x,x[length(x):1]),y=c(yl,yu[length(x):1]),col=rgb(maxColorValue=255,red=tmpCol[1],
                                                              green=tmpCol[2],blue=tmpCol[3],alpha=200),border=NA)
lines(x,y,col="black",lwd=2)

legend(x="bottomright",lty=c(1,1),lwd=c(2,10),col=c("black",rgb(maxColorValue=255,red=tmpCol[1],green=tmpCol[2],blue=tmpCol[3],
                                                                alpha=200)),legend=c("estimate","95% confidence interval"))

plot(x,moe,col="black",type="l",lwd=2,main="n=200",xlab="prevalence",ylab="margin of error")
par(xpd=T)
tmpCol<-col2rgb("steelblue")
polygon(x=c(x,x[length(x):1]),y=c(moel,moeu[length(x):1]),col=rgb(maxColorValue=255,red=tmpCol[1],
                                                                  green=tmpCol[2],
                                                                  blue=tmpCol[3],
                                                                  alpha=200),
        border=NA)
lines(x,moe,col="black",lwd=2)
legend(x="topright",lty=c(1,1),col=c("black",rgb(maxColorValue=255,red=tmpCol[1],green=tmpCol[2],blue=tmpCol[3],alpha=200)),
       lwd=c(2,10),legend=c("margin of error","95% confidence interval"))

```


## 2. Power for detecting change in seroconversion rates

### Assumptions

* Salmonella shedding prevalence (i.e. the proportion of individuals which will test positive for Salmonella in stool at a given visit): $p_{sal}=8\%$

### Method

This is quite similar to testing for a difference in SBA between 2 groups, but now we compare a binary outcome (seroconversion) between the 2 groups. Additional assumption are be around what proportion of individuals are already seroconverted at the first sample and how many more are seroconverted at the second sample -- we compute several scenarios here.

As before, we will use ready formulas (now for comparing proportions between 2 groups of unequal size), but the actual analysis in the study can be adjusted for e.g. age and other confounders.

### Results

```{r}
gr<-expand.grid(seq(1000,2500,by=100),c(0.1,0.15,0.2,0.25),c(0.2,0.3,0.4,0.5),c(1/24,0.074,0.08,0.099))

datPowSeroConv<-data.frame(
  alpha=0.05,
  prev=gr[,4],
  pDropOut=0.2,
  nTot=gr[,1],
  nTotComp=NA,
  n1=NA,
  n2=NA,
  p1=gr[,2],
  p2=gr[,3],
  power=NA
)

datPowSeroConv$nTotComp<-floor((1-datPowSeroConv$pDropOut)*datPowSeroConv$nTot)
datPowSeroConv<-datPowSeroConv[datPowSeroConv$p2/datPowSeroConv$p1<3 & datPowSeroConv$p2/datPowSeroConv$p1>1,]

datPowSeroConv$n1<-round(datPowSeroConv$prev*datPowSeroConv$nTotComp)
datPowSeroConv$n2<-round((1-datPowSeroConv$prev)*datPowSeroConv$nTotComp)
datPowSeroConv$p1p2<-factor(paste(sep="","p1 = ",datPowSeroConv$p1,", p2 = ",datPowSeroConv$p2))

for(j in 1:nrow(datPowSeroConv)){
  datPowSeroConv$power[j]<-pwr.2p2n.test(h=ES.h(p1=datPowSeroConv$p1[j],p2=datPowSeroConv$p2[j]),
                                         n1=datPowSeroConv$n1[j],n2=datPowSeroConv$n2[j])$power
}

```

Results below show that power is quite high for most scenarios, particularly when the effect of exposure results in a doubling or more of the seroconversion rate (i.e. p2 $\geq$ 2$\cdot$p1).

Note that the sample sizes go up to 2,500 on the graph below. The vertical dashed grey line indicates the 2,000 individuals that are planned to be recruited.

```{r}
ggplot(data=datPowSeroConv[datPowSeroConv$prev==0.08,],mapping=aes(x=nTot,y=power,group=p1p2,color=p1p2)) +
  geom_point(size=2) +
  geom_line(lwd=1) +
  scale_color_manual(values=colorRampPalette(brewer.pal(9,"Spectral"))(length(levels(datPowSeroConv$p1p2))),
                     name="prop. seroconverted in each group") +
  geom_hline(yintercept=0.8,lwd=2,lty=2,col="darkgrey") + 
  geom_vline(xintercept=2000,lwd=2,lty=2,col="darkgrey") +
  xlab("Number of paired samples") +
  ylab("Power") +
  ggtitle("Power curves for different combinations of 3-month seroconversion rates. Prev. of exposure = 8%, drop-out = 20%.") +
  theme_bw() +
  theme(
    legend.position = "right",
    title = element_text(face = "bold", size = 16),
    axis.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 14)
  )
```

## 3. Power for detecting change in SBA due to Salmonella exposure

### Assumptions

* Salmonella shedding prevalence (i.e. the proportion of individuals which will test positive for Salmonella in stool at a given visit): $p_{sal}=8\%$

### Method

A stool sample will only be collected at the first visit and tested for presence of Salmonella. This allows splitting the data into 2 groups: those who are shedding Salmonella at the first visit and those who do not.

As exposure will have a delayed effect on immunity / SBA, only the second SBA measurement needs to be compared (but can use the SBA measurement from the first visit to adjust for unmeasured latent factors that result in some individuals having naturally higher or lower SBA).

From a sample size / power calculation point of view this means that we can use formulas for simple, unbalanced, two-sample t-tests (technically this assumes an unadjusted analysis, but it is common to do sample size calculations this way: adjustment - if needed - should increase power).

In R we calculate this using the function `pwr.t2n.test()` from the `pwr` package.

### Results

```{r}
gr<-expand.grid(seq(1000,2500,by=100),c(0.1,0.2),c(0.4,0.6),c(1/24,0.074,0.08,0.099))

datPowSalm<-data.frame(
  alpha=0.05,
  prev=gr[,4],
  pDropOut=0.2,
  nTot=gr[,1],
  nTotComp=NA,
  n1=NA,
  n2=NA,
  delta=gr[,2],
  sd=gr[,3],
  es=NA,
  power=NA
)

datPowSalm$nTotComp<-floor((1-datPowSalm$pDropOut)*datPowSalm$nTot)
datPowSalm$n1<-round(datPowSalm$prev*datPowSalm$nTotComp)
datPowSalm$n2<-round((1-datPowSalm$prev)*datPowSalm$nTotComp)
datPowSalm$es<-round(digits=2,datPowSalm$delta/datPowSalm$sd)

for(j in 1:nrow(datPowSalm)){
  datPowSalm$power[j]<-pwr.t2n.test(n1=datPowSalm$n1[j],n2=datPowSalm$n2[j],d=datPowSalm$es[j],
                                    sig.level=datPowSalm$alpha[j])$power
}

datPowSalm$EsSd<-factor(paste(sep="","ES = ",datPowSalm$delta,", SD = ",datPowSalm$sd))

```

As the results below show, power is quite high (`r round(digits=2,100*datPowSalm$power[datPowSalm$prev==0.08 & datPowSalm$delta==0.2 & datPowSalm$nTot==2000 & datPowSalm$sd==0.4])`%) for a difference of 0.2 in SBA between the 2 groups, assuming 2,000 individuals, 20% drop-out, $\sigma_{SBA}=0.4$. If SBA has higher variance (e.g. $\sigma=0.6$ rather than $\sigma=0.4$), power will be reduced but still very high (`r round(digits=2,100*datPowSalm$power[datPowSalm$prev==0.08 & datPowSalm$delta==0.2 & datPowSalm$nTot==2000 & datPowSalm$sd==0.6])`% for $\sigma_{SBA}=0.6$).

For smaller effect sizes (difference of 0.1 in SBA) power is lower (see graph).

Note that the sample sizes go up to 2,500 on the graph below. 

```{r}
ggplot(data=datPowSalm[datPowSalm$prev==0.08,],mapping=aes(x=nTot,y=power,group=EsSd,color=EsSd)) +
  geom_point(size=2) +
  geom_line(lwd=1) +
  scale_color_manual(values=c("steelblue","greenyellow","salmon","orange"),name="effect size and std. dev.") +
  geom_hline(yintercept=0.8,lwd=1,lty=2,col="darkgrey") + 
  geom_vline(xintercept=2000,lwd=1,lty=2,col="darkgrey") +
  theme(legend.position = "right",text=element_text(size=16)) +
  xlab("Number of paired samples") +
  ylab("Power") +
  ggtitle("Power curves for different effect sizes. Prevalence of exposure = 8%, drop-out = 20%.")+
  theme_bw() +
  theme(
    title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold", size = 12),
    axis.text = element_text(size = 12)
  )

```


## 4. Sample size for longitudinal sample subset average rate of SBA change between two groups (difference in slopes)

### Community Longitudinal Cohort and Hospital Cases Cohort
This is an exploratory end-point, to provide pilot estimates for future studies due to limited existing data. We can allow a large difference in SBA change over the 12 month period on 0.75.
The same calculation will be used for both longitudinal cohorts:
1. Comparision of SBA in children with invasive NTS disease compared to healthy community controls
2. Comparison of SBA in children with known malaria exposure (MRDT+) compared to no recorded malaria exposure (MRDT-)

### Assumptions
From raw longitudinal data used in T.Nyirenda Thesis, "NATURAL IMMUNITY TO SALMONELLA IN HUMANS" (2015), using data for children 6 months - 36 months (linear increase in SBA acquisition with age)
Linear Mixed model with random effects (lmer function in lme4 package), yields the following parameters:
Variance of random intercept $\sigma^2_i = 0.299$
Variance of random slope $\sigma^2_s = 0.00148$   
Random effects residual variance $\sigma^2_e = 1.226$  
Time for follow-up 12 months (5 visits at 0,3,6,9,12 months)


### Methods
Using the "longpower" package using the "diggle.linear.power" function sample size calculation for difference in slopes between two groups, as per Diggle P.J., Heagerty P.J., Liang K., Zeger S.L. (2002) Analysis of longitudinal data. Second Edition. Oxford Statistical Science Series.

```{r}

sig2.i <- 0.299   # var of random intercept
sig2.s <- 0.00148   # var of random slope
sig2.e <- 1.226  # residual var
cov.s.i <- 0.8*sqrt(sig2.i)*sqrt(sig2.s)  # covariance of slope and intercept

cov.t <- function(t1, t2, sig2.i, sig2.s, cov.s.i){
  sig2.i + t1*t2*sig2.s + (t1+t2)*cov.s.i 
}

t = seq(0,1.5,0.25)   # study visit times
n <- length(t)
R <- outer(t, t, function(x,y){cov.t(x,y, sig2.i, sig2.s, cov.s.i)})
R <- R + diag(sig2.e, n, n)

diggle.linear.power(d=0.75, t=t, R=R, sig.level=0.05, power=0.80)   # Diggle 2002 (difference in slopes)
p50 <- diggle.linear.power(d=0.75, t=t, R=R, sig.level=0.05, power=0.50)   
p60 <- diggle.linear.power(d=0.75, t=t, R=R, sig.level=0.05, power=0.60)   
p70 <- diggle.linear.power(d=0.75, t=t, R=R, sig.level=0.05, power=0.70)   
p80 <- diggle.linear.power(d=0.75, t=t, R=R, sig.level=0.05, power=0.80)   
p90 <- diggle.linear.power(d=0.75, t=t, R=R, sig.level=0.05, power=0.90)   
p95 <- diggle.linear.power(d=0.75, t=t, R=R, sig.level=0.05, power=0.95)   
p97 <- diggle.linear.power(d=0.75, t=t, R=R, sig.level=0.05, power=0.97)

power_curve <- data.frame(
  power = c(50,60,70,80,90,95,97), 
  sample_size = c(p50$N,p60$N,p70$N,p80$N,p90$N,p95$N,p97$N)
  )

ggplot(power_curve, aes(x=sample_size, y=power)) +
  geom_point()+
  geom_line(col="blue")+
  #geom_smooth(df=2)+
  ggtitle("Power Curve for Longitudinal Cohort SBA (Change of 0.75)")+
  xlab("Sample Size")+
  ylab("Power") +
  geom_hline(yintercept=80,lwd=1,lty=2,col="red") + 
  geom_vline(xintercept=39.18,lwd=1,lty=2,col="red") +
  scale_x_continuous(breaks = seq(20,80,5))+
  theme_bw() +
  theme(
    title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold", size = 12),
    axis.text = element_text(size = 12)
  )

```

### With unbalanced design (Malaria prevalance=0.26)

```{r, results=T}
tab3 <- data.frame(cbind(
  rho = rep(c(0.0, 0.3, 0.5, 0.8), 2),
  pi1 = c(rep(0.74, 4), rep(0.3, 4))))
m <- c()
for(i in 1:nrow(tab3)){
  R <- matrix(tab3$rho[i], nrow = 4, ncol = 4)
  diag(R) <- 1
  m <- c(m, ceiling(liu.liang.linear.power(
    delta=0.75,
    u = list(u1 = rep(1, 4), # treatment
             u2 = rep(0, 4)), # control       
    v = list(v1 = rep(1, 4), v2 = rep(1, 4)), # intercept
    sigma2=1,
    Pi = c(tab3$pi1[i], 1-tab3$pi1[i]),
    R=R, alternative="two.sided",
    power=0.80)$N))
}
tab3d <- cbind(tab3, m)
flextable::flextable(tab3d)

```


